\vspace{-0.5in}{\bfseries\scshape Transfer Learning for computationally limited training of Vision Transformers on art classification tasks } \\ \vspace{1ex} \normalsize{Bachelor's Project Thesis}\vspace{-2ex}

### 2022-07-19 writing/tex/draft_conclusion.tex
Concusion concluded.

\section{Conclusion}

% TODO dit moet echt allemaal nog een stuk beter!

Seeing which is better, VTs or CNNs, has never been the goal of this paper. The underlying architecture is somewhat irrelevant, as one should always pick (or create) the model that fits the task best. What this paper, then, was interested in, is whether VTs can become a valuable alternative to CNNs under circumstances in which:
\begin{itemize}
\item small datasets are used: specifically, small enough to have been hand labelled by one person;
\item only consumer grade hardware is available.
\end{itemize}
In other words: this paper was interested in the type of machine learning one could try at home, and wanted to know if VTs can be put to good use there.

The way it tried to put them to use, is by utilising transfer learning. In the domain of art classification, it compared common CNNs and VTs to one another, for both off the shelf learning, where only the last layer of a pre-trained model is retrained on the new task, and fine-tuning, where the whole model is retrained.

The results showed that VTs are very well able to perform on par with CNNs for both off the shelf learning and fine-tuning. The most interesting architecture was the \texttt{Swin} VT, which gave best testing performance in almost all cases. Consequently, this model will be one of the firsts the author considers if he is ever faced with another computer vision task. Provided, of course, that the field does not catch up in ways that draw all models described here obsolete...

% TODO: more blabla

>>>>

\section{Conclusion}
Seeing which is better, VTs or CNNs, has never been the goal of this paper. The underlying architecture is somewhat irrelevant, as one should always pick (or create) the model that fits the task best -- regardless of whether it is a VT, CNN, or something else. What this paper, then, was interested in, is whether VTs can become a valuable alternative to CNNs under circumstances in which:
\begin{itemize}
\item small datasets are used: specifically, small enough to have been hand labelled by one person;
\item only consumer grade hardware is available.
\end{itemize}
In other words: this paper was interested in the type of machine learning one could try at home, and wanted to know if VTs could be made useful for it.

To that end, it tried to utilise transfer learning methods, as these might help overcome problems that arise from VTs' lack of an inductive bias. More concretely, models pre-trained on ImageNet1K were either used as off the shelf feature extractors, or fine-tuned to new tasks. These tasks were all taken from the \textit{Rijksmuseum Challenge} dataset \citep{mensink14icmr}, and fall within the domain of art classification.

The results demonstrated that VTs are very well able to perform on par with CNNs here. Especially the \texttt{Swin} VT showed promising results, and might be worth considering for future learning problems.

\texttt{ConvNext} was a close second in the rankings. It is interesting that this best performing CNN was inspired by VTs, while the best performing VT \texttt{Swin} took inspiration from CNNs. It shows the kind of cross-fertilization that moves the field forward, but also hints at something bigger that would be nice to conclude with.

No matter what happens. No matter if VTs, CNNs, a hybrid architecture, or even something completely different rains supreme in the coming years; it is great that VTs came about. It shakes things up, and forces researchers to look at problems from a different perspective. Perhaps that, then, will be VTs' greatest contribution of all.

### 2022-07-19 writing/tex/draft_experiments.tex
Concusion concluded.

VTs and CNNs in general are much closer together. To again take \textit{Type classification} as an example: the mean testing accuracy of all VTs combined is 91.2\% on this task, while it is 91.3\% for CNNs. Recall that section \ref{results:ots} reported a difference of 1\% in favour of VTs here, with an accuracy of 86.5\% for VTs, and 85.5\% for CNNs.

@@

\subsection{Interpretation} \label{exp:int} % TODO maybe change to Discussion instead
While most results have now been presented, it is still important to discuss them in more detail, as this leads to a more complete interpretation. This section attempts to provide such a discussion, by answering questions one might have at this moment. It will first discuss some aforementioned experiments individually, but will conclude with remarks that are more general. Topics will include: comparisons with earlier studies, potential shortcomings, and hypotheses that might explain certain observations.

@@

\paragraph{Why does \textit{Artist classification} show the best overall performance?}
This might seem counter-intuitive: one would think it is much easier to determine if something is a painting or a sculpture, than who the creator is. However, because only the top 30 classes are used, this is not necessarily true.

Figure \ref{methods:eg} shows example images for all tasks. On the one hand, it shows much heterogeneity for the \textit{Artist classification} task, since authors work in very different styles and mediums. On the other hand, it also shows difficulties for the other tasks: \textit{Material classification}, which seems to be the most challenging task, shows examples of both big differences within classes (note how the two instances of porcelain are very different), and small differences between classes (note how the porcelain plate is very similar to the faience plate).

Then there is the problem of balance. It was mentioned earlier that balanced accuracy falls behind plain accuracy the most for \textit{Material classification}, and the least for \textit{Artist classification}. At the same time, table \ref{methods:datasets} shows that it is also \textit{Material classification} which is the least balanced (gini-coefficient of 0.563), while it is \textit{Artist classification} that is the most balanced (coefficient 0.236). This suggests that balance is hurting the performance measures, so also plain accuracy, the least for \textit{Artist classification}.

@@

A small asterisk to be made here, is that VTs might benefit more from better regularisation than CNNs (see section \ref{exp:int:ft}). Still, final accuracy should not be the only concern, because ease of use will also determine to what degree VTs are a valuable alternative to CNNs. Moreover, the main finding of this paper seems to be that VTs are able to perform on par with CNNs. Slightly better performance for VTs would not exactly hurt this finding.

@@

More interesting is the question if, for example, `tiny' or `medium' sized version should have been used instead of `base' versions. On \textit{Type classification} `tiny' models were tested as well, and results were reported where appropriate. Overall, however, these results did not suggest that using `tiny' versions everywhere have would made much of a difference.

@@

>>>>

VTs and CNNs grouped together are much more evenly matched. To again take \textit{Type classification} as an example: the mean testing accuracy of all VTs combined is 91.2\% on this task, while it is 91.3\% for CNNs. Recall that section \ref{results:ots} reported a difference of 1\% in favour of VTs here, with an accuracy of 86.5\% for VTs, and 85.5\% for CNNs.

@@

\subsection{Discussion} \label{exp:int} % TODO maybe change to Interpretation instead
While most results have now been presented, it is still important to discuss them in more detail, as this leads to a more complete interpretation. This section attempts to provide such a discussion, by answering questions one might have at this moment. It will first address some of the aforementioned experiments individually, but will conclude with remarks that are more general. Topics will include: comparisons with earlier studies, potential shortcomings, and hypotheses that might explain certain observations.

@@

\paragraph{Why does \textit{Artist classification} show the best overall performance?}
This might seem counter-intuitive: one would think that it is much easier to determine if something is a painting or a sculpture, than determining who the creator is. However, because only the top 30 classes are used, this is not necessarily true.

Figure \ref{methods:eg} shows example images for all tasks. On the one hand, it shows much heterogeneity for the \textit{Artist classification} task, with artists working in very different styles and mediums. On the other hand, it also shows difficulties for the other tasks: \textit{Material classification}, which seems to be the most challenging task, shows examples of both big differences within classes (note how the two instances of porcelain are very different), and small differences between classes (note how the porcelain plate is very similar to the faience plate).

Then there is the problem of balance. It was mentioned earlier that balanced accuracy falls behind plain accuracy the most for \textit{Material classification}, and the least for \textit{Artist classification}. At the same time, table \ref{methods:datasets} shows that it is also \textit{Material classification} which is the least balanced (gini-coefficient of 0.563), while it is \textit{Artist classification} that is the most balanced (coefficient 0.236). This suggests that balance is hurting performance the least for \textit{Artist classification} -- and the most for \textit{Material classification}.

@@

A small asterisk to be made here, is that VTs might benefit more from better regularisation than CNNs (see section \ref{exp:int:ft}). Still, final accuracy should not be the only concern, because ease of use (i.e. effort needed to find the right hyperparameters) will also determine to what degree VTs are a valuable alternative to CNNs. Moreover, the main finding of this paper seems to be that VTs are able to perform on par with CNNs. Slightly better performance for VTs would not exactly hurt this finding.

@@

It is also interesting to wonder if, for example, `tiny' or `medium' sized version should have been used instead of `base' versions. All available `tiny' models were tested as well, and results have been reported where appropriate (e.g. section \ref{exp:int:ft}). Overall, however, these results did not suggest that using `tiny' versions would have made much of a difference for the rankings: `tiny' \texttt{Swin} and \texttt{ConvNext}, for example, remained superior under that circumstance.

@@

After OTS learning, results are slightly better for VTs, while after FT, results are almost the same for CNNs and VTs. With that said, it seems, VTs are very much able to perform on par with CNNs on the described art classification problems. Especially \texttt{Swin} and \texttt{DeiT} showed promising performance, and together with \texttt{ConvNext}, might be interesting models to consider.

Of course, final accuracy is not the full story, since time/accuracy trade-offs, for example, might also determine which model is preferred. Figure \ref{results:img:ots_vs_ft_type} plots \textit{Type classification} accuracies of tables \ref{results:tab:ots} and \ref{results:tab:ft} against training time per epoch. Total training time was not chosen instead, as this measure depends much more on selected hyperparameters.

The best overall performing model is a fine-tuned VT (\texttt{Swin}). At the same time, it is a CNN that shows the best time/accuracy trade-offs, because it is able to perform better than any OTS trained model, while still taking less time per epoch than any OTS trained VT. The particular CNN this regards, is \texttt{ResNet50}. Recall that this model only had 25.6 million parameters, while most VTs had around 86 million.

All `tiny' VTs are around the same size as \texttt{ResNet50}, and using these would have changed the situation here completely. As such, taking figure \ref{results:img:ots_vs_ft_type} with a grain of salt, would be appropriate.

Still, it seemed fitting to conclude with this figure, as it neatly summarises all the main findings done in this section.

### 2022-07-19 writing/tex/draft_introduction.tex
Concusion concluded.

Another ``shortcomming'' of \citeauthor{matsoukas2021time} is that it merely examined whether VTs can use TL to outperform CNNs in a domain that happens to be characterized by limited-sized datasets, but that it did not actively focus on the extent to which VTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make VTs applicable for smaller datasets. This is not necessarily related to TL as defined in this paper, however.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows VTs to dominate CNNs in limited compute/data environments
\\\\
The current study will examine to what extent TL techniques (both FT and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempts to find a methodology that is recommended in these circumstances. It will, however, restrict itself to art classification tasks. Several experiments will be performed, of which results will be analysed both quantitatively as well as qualitatively.

>>>>

Another `shortcomming' of \citeauthor{matsoukas2021time} is that it merely examined whether VTs can use TL to outperform CNNs in a domain that happens to be characterized by limited-sized datasets, but that it did not actively focus on the extent to which VTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make VTs applicable for smaller datasets. This is not necessarily related to TL as defined in this paper, however.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows VTs to dominate CNNs in limited compute/data environments
\\\\
The current study will examine to what extent TL techniques (both FT and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempts to find a methodology that is recommended in these circumstances. It will, however, restrict itself to art classification tasks. Several experiments will be performed, of which results will be analysed both quantitatively as well as qualitatively.
